### 2019-09-29 ch7_incgraph/content.tex
minor changes

The differences in graphlet counts $\Delta_{G, G'}$ are calculated by iterating recursively over the neighbours surrounding each of the modified edges (See \nameref{code_incgraph}). Several strategies are used in order to calculate $\Delta_{G,G'}$ as efficiently as possible (Fig~\ref{fig_efficient}).
(A) The delta matrix is calculated separately for each modified edge. Since the edge already contains two out of five nodes and any modified graphlet is a connected subgraph, the neighbourhood of this edge only needs to be explored up to depth 3 in order to iterate over all modified graphlets. 
(B) We propose a lookup table to look up the graphlet index of each node of a given subgraph. By weighting each possible edge in a 5-node graphlet, the sum of the edges of a subgraph can be used to easily look up the corresponding graphlet index.
(C) During the recursive iteration of the neighbourhood, the same combination of nodes is never visited twice.
(D) As the network can be relatively large, the adjacency matrix is binary compressed in order to save memory. One integer requires 4 bytes and contains the adjacency boolean values of 32 edges, whereas otherwise 32 booleans would require 32 bytes. For example, 1GB of memory is large enough to store a compressed adjacency matrix of 92681 nodes. If the network contains too many nodes to fit a compressed adjacency matrix into the memory, a list of sets containing each node's neighbours is used instead.

@@

The network models were based on three static network models: Barab\'asi-Albert~\cite{albert_statisticalmechanicscomplex_2002}, Erd\H{o}s--R\'enyi~\cite{erdos_randomgraphs_1959}, and Geometric~\cite{appel_minimumvertexdegree_1997}. Pseudo code for these random evolving network models can be found in \nameref{code_ba}, \nameref{code_er}, and \nameref{code_geo} respectively.
Each model generates an initial network according to the static network model it is based on, and a list of network modifications (removing an edge from or adding an edge to the network). These network modifications are made such that at any given time point in the evolving network, it is likely that the network at its current state could have been generated by the original static network model.

@@

Comparing the execution time of IncGraph to the h-index of the networks indicates that the amortised time of IncGraph could be $O(h^3)$ (\nameref{sfig_hindex}). This is in line with the amortised times $O(h)$ and $O(h^2)$ of the algorithm described by Eppstein et al.\cite{eppstein_extendeddynamicsubgraph_2012} for counting three-size and four-size subgraphs respectively.

@@

\section*{Supporting information}

%% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{S1 Pseudocode.}
\label{code_incgraph}
{\bf IncGraph calculates $\Delta_{G,G'}$ using a strict branch-and-bound strategy.}

\paragraph*{S2 Pseudocode.}
\label{code_ba}
{\bf Pseudo code for generating an evolving Barab\'asi-Albert (BA) network.} It first generates a BA network, and then generates $o$ operations such that at any time point, the network is or very closely resembles a BA network.

\paragraph*{S3 Pseudocode.}
\label{code_er}
{\bf Pseudo code for generating an evolving Erd\H{o}s--R\'enyi (ER) network.} It first generates an ER network, and then generates $o$ operations such that at any time point, the network is or very closely resembles an ER network.

\paragraph*{S4 Pseudocode.}
\label{code_geo}
{\bf Pseudo code for generating an evolving geometric network.} It first generates a geometric network, and then generates $o$ operations such that at any time point, the network is or very closely resembles a geometric network.

\paragraph*{S1 Fig.}
\label{sfig_hindex}
{\bf Empirical measurements show a strong relation between the execution time of IncGraph and the h-index cubed of the network it was applied to.} This is in line with the findings by Eppstein et al., where counting 3-size subgraphs has an amortised time of $O(h)$ and counting 4-size subgraphs has an amortised time of $O(h^2)$.

>>>>

The differences in graphlet counts $\Delta_{G, G'}$ are calculated by iterating recursively over the neighbours surrounding each of the modified edges (Fig~\ref{code_incgraph}). Several strategies are used in order to calculate $\Delta_{G,G'}$ as efficiently as possible (Fig~\ref{fig_efficient}).
(A) The delta matrix is calculated separately for each modified edge. Since the edge already contains two out of five nodes and any modified graphlet is a connected subgraph, the neighbourhood of this edge only needs to be explored up to depth 3 in order to iterate over all modified graphlets. 
(B) We propose a lookup table to look up the graphlet index of each node of a given subgraph. By weighting each possible edge in a 5-node graphlet, the sum of the edges of a subgraph can be used to easily look up the corresponding graphlet index.
(C) During the recursive iteration of the neighbourhood, the same combination of nodes is never visited twice.
(D) As the network can be relatively large, the adjacency matrix is binary compressed in order to save memory. One integer requires 4 bytes and contains the adjacency boolean values of 32 edges, whereas otherwise 32 booleans would require 32 bytes. For example, 1GB of memory is large enough to store a compressed adjacency matrix of 92681 nodes. If the network contains too many nodes to fit a compressed adjacency matrix into the memory, a list of sets containing each node's neighbours is used instead.

@@

The network models were based on three static network models: Barab\'asi-Albert~\cite{albert_statisticalmechanicscomplex_2002}, Erd\H{o}s--R\'enyi~\cite{erdos_randomgraphs_1959}, and Geometric~\cite{appel_minimumvertexdegree_1997}. These models were adapted to generate evolving networks instead (Figs~\ref{code_ba}, \ref{code_er}, and \ref{code_geo}).
Each model generates an initial network according to the static network model it is based on, and a list of network modifications (removing an edge from or adding an edge to the network). These network modifications are made such that at any given time point in the evolving network, it is likely that the network at its current state could have been generated by the original static network model.

@@

Comparing the execution time of IncGraph to the h-index of the networks indicates that the amortised time of IncGraph could be $O(h^3)$ (Fig~\ref{sfig_hindex}). This is in line with the amortised times $O(h)$ and $O(h^2)$ of the algorithm described by Eppstein et al.\cite{eppstein_extendeddynamicsubgraph_2012} for counting three-size and four-size subgraphs respectively.

@@

\section{Supporting information}

\newcommand{\pluseq}{\mathrel{+}=}
\newcommand{\asteq}{\mathrel{*}=}

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.8\linewidth]{fig/figureS1.pdf} 
	\caption{
		\textbf{Empirical measurements show a strong relation between the execution time of IncGraph and the h-index cubed of the network it was applied to.} This is in line with the findings by Eppstein et al., where counting 3-size subgraphs has an amortised time of $O(h)$ and counting 4-size subgraphs has an amortised time of $O(h^2)$.
	}
	\label{sfig_hindex}
\end{figure}

\begin{figure}[htb!]
	\begin{algorithmic}
		\Function{CalculateDelta}{Network $G'$, Node $n_0$, Node $n_1$}
		\State \Comment{Assuming an edge $(n_0, n_1)$ has just been added or removed from $G'$}
		\State $\Delta^- \gets \Call{Matrix}{\Call{NumNodes}{G}, 73}$ \Comment{For storing the orbit counts of the removed graphlets}
		\State $\Delta^+ \gets \Call{Matrix}{\Call{NumNodes}{G}, 73}$ \Comment{For storing the orbit counts of the added graphlets}
		\State $B_0 = \{n_0, n_1\}$ \Comment{A blacklist of nodes not to visit anymore}
		\State $e = (n_0, n_1)$ \Comment{Different name for the edge}
		\If{$e \in N'$}
		\State $(m^-, m^+) = (0, 1)$ \Comment{In other words, $m^- = 1$ iff $e \in N$ and $m^+ = 1$ iff $e \in N'$}
		\Else
		\State $(m^-, m^+) = (1, 0)$
		\EndIf
		\State $x_0 = 0$
		\State $(\Delta^+, \Delta^-) = \Call{CountOrbits}{\Delta^+, \Delta^-, (n_0, n_1), x_0, m^-, m^+}$ \Comment{Update delta matrices for current nodes}

		%\For{$n_2 \in \Call{Neighbours}{S_0}$}
		\For{$n_2 \in \bigcup_{i \in \{0, 1\}}\Call{Neighbours}{n_i}$}
		\If{$n_2 \not\in B_0$}
		\State $B_0 = \{n_2\} \bigcup B_0$ \Comment{Add $n_2$ to blacklist $B_0$}
		\State $x_1 = x_0 + \Call{W}{G', (n_0, n_2), 1} + \Call{W}{G', (n_1, n_2), 2}$ \Comment{Calculate edge weights for current nodes}
		\State $(\Delta^+, \Delta^-) = \Call{CountOrbits}{\Delta^+, \Delta^-, (n_0, n_1, n_2), x_1, m^-, m^+}$
		\State $B_1 = B_0$ \Comment{Make a copy for the next iteration depth}
		\For{$n_3 \in \bigcup_{i \in \{0, 1, 2\}}\Call{Neighbours}{n_i}$}
		\If{$n_3 \not\in B_1$}
		\State $B_1 = \{n_3\} \bigcup B_1$
		\State $x_2 = x_1 + \Call{W}{G', (n_0, n_3), 3} + \Call{W}{G', (n_1, n_3), 4} + \Call{W}{G', (n_2, n_3), 5}$
		\State $(\Delta^+, \Delta^-) = \Call{CountOrbits}{\Delta^+, \Delta^-, (n_0, n_1, n_2, n_3), x_2, m^-, m^+}$
		\State $B_2 = B_1$

		\For{$n_4 \in \bigcup_{i \in \{0, 1, 2, 3\}}\Call{Neighbours}{n_i}$}
		\If{$n_4 \not\in B_2$}
		\State $B_2 = \{n_4\} \bigcup B_2$
		\State $x_3 = x_2 + \Sigma_{i \in \{0, 1, 2, 3\}}\Call{W}{G', (n_i, n_4), i+6}$
		\State $(\Delta^+, \Delta^-) = \Call{CountOrbits}{\Delta^+, \Delta^-, (n_0, n_1, n_2, n_3, n_4), x_3, m^-, m^+}$
		\EndIf
		\EndFor
		\EndIf
		\EndFor
		\EndIf
		\EndFor
		\EndFunction
		\Function{CountOrbits}{$\Delta^+$, $\Delta^-$, Nodes $S$, Edgeweights $x$, Modifier $m^-$, Modifier $m^+$}
		\State $L^- = L[x + m^-]$ \Comment{Look up the orbits of the subgraph induced by $S$ in $N$}
		\State $\Delta^-[S, L^-] \pluseq 1$ \Comment{Increment orbit counts of nodes $S$ at positions $L^-$ in $\Delta^-$}
		\State $L^+ = L[x + m^+]$ \Comment{Look up the orbits of the subgraph induced by $S$ in $N'$}
		\State $\Delta^+[S, L^+] \pluseq 1$ \Comment{Increment orbit counts of nodes $S$ at positions $L^+$ in $\Delta^+$}
		\State \Return $(\Delta^-, \Delta^+)$
		\EndFunction
		\Function{W}{Network $G$, Edge $e$, Exponent $i$}
		\State \Return $e \in \Call{Edges}{G} \ ?\  2^i : 0$ \Comment{Return $2^i$ if $G$ contains edge $e$}
		\EndFunction
	\end{algorithmic}
	\caption{
		\textbf{Pseudocode of IncGraph.} IncGraph calculates $\Delta_{G,G'}$ using a strict branch-and-bound strategy.
	}
	\label{code_incgraph}
\end{figure}

\begin{figure}[htb!]	
	\begin{algorithmic}
		\Function{BarabasiAlbert}{number of nodes $n$, degree $d$, \par\hspace*{3.72cm} number of operations $o$, offset exponent $x=1$}
		\State $D \gets \Call{Rep}{0, n}$ \Comment{All nodes start with 0 degree}
		\State $T \gets \Call{Rep}{\{\}, n}$ \Comment{List of targets of each node}
		\State $S \gets \Call{Rep}{\{\}, n}$ \Comment{List of sources of each node}
		\State $O \gets (\,)$ \Comment{Variable to store the generated operations in}

		\Procedure{AddEdge}{node $i$, node $j$, add as operation $b$}
		\State $D[i] \gets D[i] + 1$;\ $D[j] \gets D[j] + 1$ \Comment{Update degrees}
		\State $T[i] \gets T[i] \cup \{j\}$;\ $S[j] \gets S[j] \cup \{i\}$ \Comment{Update targets and sources}
		\If{$b$}
		\State $O \gets O + (\textrm{ADD}, (i, j))$
		\EndIf
		\EndProcedure
		\Procedure{RemoveEdge}{node $i$, node $j$, add as operation $b$}
		\State $D[i] \gets D[i] - 1$;\ $D[j] \gets D[j] - 1$ \Comment{Update degrees}
		\State $T[i] \gets T[i] \setminus \{j\}$;\ $S[j] \gets S[j] \setminus \{i\}$ \Comment{Update targets and sources}
		\If{$b$}
		\State $O \gets O + (\textrm{REM}, (i, j))$
		\EndIf
		\EndProcedure
		\Procedure{AddNode}{node $i$, add as operations $b$}
		\State $C \gets \{j \,|\, 0 \leq j < i \wedge j \not\in S[i]\}$ \Comment{Select candidate neighbours}
		\State $W \gets (D[C] \,/\, \Sigma D[C])^x$ \Comment{Calculate weights for preferred attachment}
		\State $X \gets $ sample $d$ neighbours from $C$ with weights $W$
		\For{$j \in X$}
		\State \Call{AddEdge}{$i$, $j$, $b$}
		\EndFor
		\EndProcedure
		\Procedure{RemoveNode}{node $i$, add as operations $b$}
		\While{$|T[i]| > 0$}
		\State $j \gets \Call{Head}{T[i], 1}$
		\State \Call{RemoveEdge}{$i$, $j$, $b$}
		\EndWhile
		\EndProcedure
		\For{$i \in \{1 \,..\, m\}$}
		\For{$j \in \{0 \,..\, i-1\}$}
		\State \Call{AddEdge}{$i$, $j$, false} \Comment{Start with $m+1$ complete graph}
		\EndFor
		\EndFor
		\For{$i \in \{m+1\,..\,n-1\}$}
		\State \Call{AddNode}{$i$, false} \Comment{Add the rest of the nodes}
		\EndFor
		\State $G_0 \gets \{ (i, j) \,|\, i \in \{0\,..\,n-1\} \wedge j \in N[i]\}$ \Comment{Initial network}
		\While{$|O| < o$} \Comment{Modify network until $O$ is sufficiently large}
		\State $i \gets$ Sample 1 index from $\{0 \,..\, n-1\}$
		\State \Call{RemoveNode}{$i$, true}
		\State \Call{AddNode}{$i$, true}
		\EndWhile
		\State $O \gets \Call{Head}{O, o}$ 
		\State \Return $(N_0, O)$ \Comment{Return the initial network and $o$ operations}
		\EndFunction
	\end{algorithmic}
	\caption{
		\textbf{Pseudo code for generating an evolving Barab\'asi-Albert (BA) network.} It first generates a BA network, and then generates $o$ operations such that at any time point, the network is or very closely resembles a BA network.
	}
	\label{code_ba}
\end{figure}

\begin{figure}[htb!]
	\begin{algorithmic}
		\Function{ErdosRenyi}{number of nodes $n$, number of edges $e$, number of operations $o$}
		\State $P \gets \{(i, j) \,|\, i \in \{1 \,..\, n-1\} \wedge j \in \{0 \,..\, i-1\}\}$ \Comment{All possible interactions}
		\State $N_0 \gets$ Sample $e$ edges from $P$
		\State $N_c \gets N_0$
		\While{$|O|<o$}
		\State $e_a \gets $ Sample 1 edge from $P \setminus N_c$
		\State $e_r \gets $ Sample 1 edge from $N_c$
		\State $N_c \gets (N_c \setminus e_r) \cup \{e_a\}$
		\State $O \gets O + ((\textrm{ADD}, e_a), (\textrm{REM}, e_r))$
		\EndWhile
		\State $O \gets \Call{Head}{O, o}$ 
		\State \Return $(N_0, O)$ \Comment{Return the initial geometric network and $o$ operations}
		\EndFunction
	\end{algorithmic}
	\caption{
		\textbf{Pseudo code for generating an evolving Erd\H{o}s--R\'enyi (ER) network.} It first generates an ER network, and then generates $o$ operations such that at any time point, the network is or very closely resembles an ER network.
	}
	\label{code_er}
\end{figure}

\begin{figure}[htb!]
	\begin{algorithmic}
		\Function{Geometric}{number of nodes $n$, number of edges $e$, \par\hspace*{2.77cm} number of operations $o$, number of dimensions $d=3$}
		\State $P \gets$ Sample $n$ points from a multivariate continuous uniform distribution $U_{d}((0,1)^{d})$
		\State $D \gets$ Calculate distance matrix from $P$
		\State $N_0 \gets \Call{Head}{\Call{ArgSort}{\Call{LowerTriangle}{D}}, e}$ \Comment{Initial network}
		\State $N_p \gets N_0$
		\State $O \gets (\,)$ \Comment{Variable to store the generated operations in}
		\While{$|O| < o$} \Comment{Modify network until $O$ is sufficiently large}
		\State $i \gets $ Sample 1 index from $\{0\,..\,n-1\}$
		\State $P[i] \gets$ Sample 1 point from a $U_{d}((0,1)^{d})$ \Comment{Give node $i$ a new location}
		\State $D[i,] \gets D[,i] \gets $ Calculate new distances between node $i$ and all other nodes
		\State $N_c \gets \Call{Head}{\Call{ArgSort}{\Call{LowerTriangle}{D}}, e}$ \Comment{New network}
		\State $O_a \gets \{(\textrm{ADD}, e) \,|\, e \not\in E(N_p) \wedge e \in E(N_c) \}$ \Comment{Gather added edges}
		\State $O_r \gets \{(\textrm{REM}, e) \,|\, e \in E(N_p) \wedge e \not\in E(N_c) \}$ \Comment{Gather removed edges}
		\State $O \gets O + \Call{Shuffle}{O_a + O_r}$ \Comment{Append new operations to $O$}
		\State $N_p \gets N_c$
		\EndWhile
		\State $O \gets \Call{Head}{O, o}$ 
		\State \Return $(N_0, O)$ \Comment{Return the initial geometric network and $o$ operations}
		\EndFunction
	\end{algorithmic}
	\caption{
		\textbf{Pseudo code for generating an evolving geometric network.} It first generates a geometric network, and then generates $o$ operations such that at any time point, the network is or very closely resembles a geometric network.
	}
	\label{code_geo}
\end{figure}

### 2019-09-29 ch8_discussion/content.tex
added guidelines for benchmarking
minor changes

\blindtext

\section{Overview and conclusions of the presented work}

% notice that normalisation was absent from this dissertation.
%Computer scientists should proceed with caution in developing new software, however, as the results produced should not only be convincing and easy to interpret, but the software should be robust and generate sufficiently accurate models of the underlying system.
%
%Bit too excited -- false positives, poor accuracy, scalability issues, poor software quality.

\section{Future research directions}

>>>>

% DISCUSSION CONTENT

This chapter reflects on the impact of this work on the field, referring back to the research objectives that were defined in Chapter \ref{chap:introduction}.

\section{dyngen: benchmarking with \textit{in silico} single cells}
We developed a single-cell simulator that can be used to simulate types of single-cell omics data for which the technology does not yet exist (Chapter~\ref{chap:dyngen}). The \textit{in silico} single cell data can help kick-start emerging domains with low data availability more safely by making novel computational tools more reproducible. Indeed, the simulator has already been used to evaluate trajectory inference\cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment\cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and single cell network inference\cite{pratapa_benchmarkingalgorithmsgene_2019} methods.

\section{dynbenchmark: A comparison of single-cell trajectory inference methods}
Using this simulator and a collection of real datasets, we performed a comparison of 45 TI methods (Chapter~\ref{chap:dynbenchmark}). 
Our contributions include writing software to run 45 different error-prone TI methods with a common interface, downloading and processing hundreds of single cell datasets, and developing novel metrics for comparing ground truth and predicted trajectories.

We constructed a set of guidelines for end-users to help choose a TI method that is accurate, robust and fit for their application. 
Since such guidelines were hitherto lacking, they are commonly disseminated in manuscripts \cite{lafzi_tutorialguidelinesexperimental_2018,luecken_currentbestpractices_2019}, courses \cite{kiselev_analysissinglecell_2019,martens_analysissinglecell_2019}, and slides shown during keynote caffeine refuelling sessions \cite{hemberg_coffeebreakanalysis_2019}.

We made our pipeline, datasets, metrics, and containerised wrappers of TI methods publicly available for developers to use. Despite our best efforts, we do not observe an increase in developers performing quantitative benchmarks. We hypothesise causal reasons for this phenomenon and provide solutions in order to spur TI developers to perform more self-assessments (Chapter~\ref{chap:selfassessment}).

Over the course of four years, we developed a lot of practical experience in benchmarking complex computational tools. We summarise this experience in a set of essential guidelines for benchmarking computational tools (Chapter~\ref{chap:guidelines}).

%	\item Already outdated when the manuscript was published online
%	\item Update benchmark with more TI methods, newer (and larger) datasets, perform parameter optimisation on methods
%	\item Include RNA velocity as inputs

\section{dyno: A toolkit for inferring and interpreting trajectories}
% spinoff from dynbenchmark
% not yet published, yet a lot of activity on github
% invest time in publishing all of the subpackages on github

\section{SCORPIUS: Fast, accurate, and robust single-cell pseudotime}

%\item We introduce a novel TI method specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}). Despite linear TI being the most simple but commonly used form of trajectory inference, the benchmark demonstrated that most TI methods are not capable of producing accurate models of linear datasets.

%\section{SCORPIUS discussion}
%\begin{itemize}
%	\item Extension to inferring non-linear trajectories, i.e. with principal graphs or GNG
%\end{itemize}

\section{bred: Inferring single cell regulatory networks}

%\item We invent a new type of NI method capable of inferring the GRN of individual cells (Chapter~\ref{chap:bred}). We demonstrate this <yadeyade .. fill in when the chapter is actually written.>

\section{incgraph: Optimising regulatory networks}

%\item Every NI method has certain topological biases. We provide a tool for analysing the topological properties of large, evolving networks and use this to iteratively optimise GRN predictions (Chapter~\ref{chap:incgraph}).
